Evolution of ORM Systems: Formal Foundations

### 2013-01-19 doc/description.tex
textreview

There are three main goals of our evolutionary framework: 1) provide one source of evolution definition for both: code and database 2) provide a tool, which will be able to preserve stored data 3) minimize the manual work during data evolution in a software.

@@

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd Software \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}
The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
 	s \lhd Software
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Value \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \lhd Row*, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\paragraph{Mapping of Application} The goal of ORM is usually to determine how an instance is stored in the database. The ORM maps the structure of application classes to database schema. The assumption is a software is evolved continuously and evolutionary transformations produces consistent software, nevertheless to verify consistency of a software we need mapping between an existing application structure and database schema. 
\begin{align*}
	\mathbf{\rho_{app}} : Application \rightarrow Database\\
	\rho_{app}(a) = \inference{state(a) = Consistent a}{\begin{gathered}
	db(\varnothing) \bullet addTable(classes(a)) \\ 
	\bullet (\forall c \in classes(a), p \in properties(c) : addProperty(name(c), p) \\
	\bullet (\forall c \in classes(a), e \in associations(c) : addAssociaiton(name(c), e) 
\end{gathered}}
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		\inference{c = \varnothing}{\varnothing}\\\\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \\ \\
		\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))}
 \end{cases}\\
 c \lhd Class
\end{align*}

@@

There can occur name collision during mapping in case there are more classes with property representing collection. To avoid that the mapping changes the name of the table representing a collection.
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
	p \lhd Property
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(t)+name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(t)+name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
 	p \lhd Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties and the same solution is used to avoid name collisions.
\begin{align*}
	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
	\rho_a(a) = \begin{cases}
		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}\\
	 a \lhd Association
\end{align*}
\begin{align*} 
	\mathbf{\rho_a} : Association* \rightarrow Table* \\
	\rho_a(a) = \begin{cases}
 		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{  startCardinality > 1 \\ \wedge \exists t \in tables(d) : name(owningClass(p)) = name(t) \\ \wedge \exists e \in tables(d) : name(reference(a)) = name(e)
  		}{\begin{gathered}  
		 table(name(t)+name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	\end{cases}\\
	 a \lhd Association
\end{align*}

@@

\subsection{Application Transformation}
This section provide transformations for manipulating the application layer of a software - the $AppT$ set. The set defines common operation for creation, alternation and decomposition of a model (or code) .

@@

\paragraph{Rename Property and Association} Both renaming operations can change a name of the property, association respectively if there are no name collisions:
\begin{align*}
	\mathbf{renProp} : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n
	}{\begin{gathered}
		p.name = l_n 
	\end{gathered}}
\end{align*}
\begin{align*}
	\mathbf{renAssoc} : Label \times Label \times Label \times Application \rightarrow Application \\
	renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n
	}{\begin{gathered}
		b.name = l_n 
	\end{gathered}}
\end{align*}

\paragraph{Rename Class} When a class is renamed, the name collisions has to be verified and then all referencing associations has to be updated.
\begin{align*}
	\mathbf{renClass} :  Label \times Label \times Application \rightarrow Application \\
	renClass(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_o \\ \wedge \forall e \in classes(a) : name(e) \neq l_n
	}{\begin{gathered}
		c.name = l_n \wedge \forall e \in classes(a) \\ \wedge \forall (f \in associations(e) \wedge reference(f) = l_o ) :\\ f.reference = l_n 
	\end{gathered}}
\end{align*}

@@

\paragraph{Compare} The need to compare two tables during the evolution results into the comparing operator $\cong$. The operator has to be defined for columns and foreign keys too.
\begin{align*}
	\mathbf{\cong} : Table \times Table \rightarrow Boolean \\
	t_1 \cong t_2 = \begin{cases}
 		\inference{|columns(t_1)| = |columns(t_2)| \\ \wedge |foreignKeys(t_1)| = |foreignKeys(t_2)| \\ \wedge \forall c \in columns(t_1) \exists e \in columns(t_2) : c = e \\ \wedge \forall f \in foreignKeys(t_1) \exists g \in foreignKeys(t_2) : f = g }{true}\\\\
 false
 \end{cases}\\
 t_1 \lhd Table, t_2 \lhd Table
\end{align*}
\begin{align*}
	\mathbf{\cong} : Column \times Column \rightarrow Boolean \\
	c_1 \cong c_2 = \begin{cases}
 		\inference{name(c_1) = name(c_2) \\ \wedge type(c_1) = type(c_2) \\ \wedge constraints(c_1) = comstraints(c_2)}{true}
 		\\ \\
		false
 \end{cases}\\
 c_1 \lhd Column, c_2 \lhd Column
\end{align*}
\begin{align*}
	\mathbf{\cong} : ForeignKey \times ForeignKey \rightarrow Boolean \\
	f_1 \cong f_2 = \begin{cases}
 		\inference{name(f_1) = name(f_2) \\ \wedge reference(f_1) = reference(f_2) \\ \wedge constraints(f_1) = comstraints(f_2}{true}
 		\\ \\
		false
 \end{cases}
\end{align*}

\paragraph{Find pair} The function returns a pair corresponding to given column (or key) name in concrete row.
\begin{align*}
\mathbf{findPair} : Label \times Row \rightarrow Pair \\
	findPair(l, r) = \begin{cases}
		\inference{\exists p \in pairs(r) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
		l \in Label, r \lhd Row
\end{align*}

\paragraph{Add Column} A new column can be add into a table if there are no name collisions. If the column has the $NOTNULL$ constraint, then the default value has to be added to all rows in the table.
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \not \in constraints(c) }{columns(t) \cup c}\\ \\
		(\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \in constraints(c) }{columns(t) \cup c})\\ \bullet addValueToTable(l, [name(c), defVal(c)]) \\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}

\paragraph{Alter Column Name}
\begin{align*}
\mathbf{alterColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) =
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
 \\
l_c, l_o, l_n \in Label, d \lhd Database
\end{align*}

\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{d \bullet addTable(head(t)) \bullet addTable(tail(t))}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge \forall f \in foreignKeys(t) : \\ s\exists e \in tables(d) \wedge name(e) = reference(f) }{tables(d) \cup t}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}

@@

\paragraph{Insert row} The function allow to insert a row(s) into a database.
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{d \bullet insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) \\ \wedge \forall e \in tableData(t, d) : pkValue(e) \neq pkValue(r)}{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}

>>>>

There are three main goals of our evolutionary framework: 1) provide one source of evolution definition for both: code and database 2) provide a tool, which will be able to preserve stored data and information.

@@

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd Software \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
 	s \lhd Software
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model. The notation: $Type*$ is used to define an ordered set of values of concrete type (e.g. $C \lhd Class*$ means c is an ordered set of classes).

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}
 \end{cases}\\
 y \lhd Software \vee y \lhd Application \vee y \lhd Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant.

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Value \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		%\inference{r = \varnothing}{\varnothing} \\
 		%\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \lhd Row*, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\paragraph{Mapping of Application} The goal of ORM is usually to determine how an instance is stored in the database. The ORM maps the structure of application classes to database schema. The assumption is a software is evolved continuously and evolutionary transformations produces consistent software, nevertheless to verify consistency of a software we need mapping between an existing application structure and database schema. 
\begin{align*}
	\mathbf{\rho_{app}} : Application \rightarrow Database\\
	\rho_{app}(a) = \inference{state(a) = Consistent \; a}{\begin{gathered}
	db(\varnothing) \bullet addTable(classes(a)) \\ 
	\bullet (\forall c \in classes(a), p \in properties(c) : addProperty(name(c), p) \\
	\bullet (\forall c \in classes(a), e \in associations(c) : addAssociaiton(name(c), e) 
\end{gathered}}
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		%\inference{c = \varnothing}{\varnothing}\\\\
		%\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))} \\ \\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \end{cases}   \\ 	
 c \lhd Class
\end{align*}

@@

There can occur name collision during mapping in case there are more classes with property representing collection. To avoid that the mapping changes the name of the table representing a collection.
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
	p \lhd Property
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(t)+name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(t)+name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
 	p \lhd Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties and the same solution is used to avoid name collisions.
\begin{align*}
	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
	\rho_a(a) = \begin{cases}
		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}\\
	 a \lhd Association
\end{align*}
\begin{align*} 
	\mathbf{\rho_a} : Association* \rightarrow Table* \\
	\rho_a(a) = \begin{cases}
 		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{  startCardinality > 1 \\ \wedge \exists t \in tables(d) : name(owningClass(p)) = name(t) \\ \wedge \exists e \in tables(d) : name(reference(a)) = name(e)
  		}{\begin{gathered}  
		 table(name(t)+name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	\end{cases}\\
	 a \lhd Association
\end{align*}

@@

\subsection{Application Transformation}
This section provide transformations for manipulating the application layer of a software - the $AppT$ set. The set defines common operation for creation, alternation and decomposition of a model (or code).

@@

\paragraph{Rename Property and Association} Both renaming operations can change a name of the property, association respectively if there are no name collisions:
\begin{align*}
	\mathbf{renProp} : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n
	}{\begin{gathered}
		p.name = l_n 
	\end{gathered}}\\
	l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}
\begin{align*}
	\mathbf{renAssoc} : Label \times Label \times Label \times Application \rightarrow Application \\
	renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n
	}{\begin{gathered}
		b.name = l_n 
	\end{gathered}}
	l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}

\paragraph{Rename Class} When a class is renamed, the name collisions has to be verified and then all referencing associations has to be updated.
\begin{align*}
	\mathbf{renClass} :  Label \times Label \times Application \rightarrow Application \\
	renClass(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_o \\ \wedge \forall e \in classes(a) : name(e) \neq l_n
	}{\begin{gathered}
		name(c) = l_n \wedge \forall e \in classes(a) \\ \wedge \forall (f \in associations(e) \wedge reference(f) = l_o ) :\\ f.reference = l_n 
	\end{gathered}}
\end{align*}

@@

\paragraph{Comparison of tables} The need to compare two tables during the evolution results into the comparing operator $\cong$. The operator has to be defined for columns and foreign keys too.
\begin{align*}
	\mathbf{\cong} : Table \times Table \rightarrow Boolean \\
	t_1 \cong t_2 = \begin{cases}
 		\inference{|columns(t_1)| = |columns(t_2)| \\ \wedge |foreignKeys(t_1)| = |foreignKeys(t_2)| \\ \wedge \forall c \in columns(t_1) \exists e \in columns(t_2) : c = e \\ \wedge \forall f \in foreignKeys(t_1) \exists g \in foreignKeys(t_2) : f = g }{true}\\\\
 false
 \end{cases}\\
 t_1 \lhd Table, t_2 \lhd Table
\end{align*}
\begin{align*}
	\mathbf{\cong} : Column \times Column \rightarrow Boolean \\
	c_1 \cong c_2 = \begin{cases}
 		\inference{name(c_1) = name(c_2) \\ \wedge type(c_1) = type(c_2) \\ \wedge constraints(c_1) = comstraints(c_2)}{true}
 		\\ \\
		false
 \end{cases}\\
 c_1 \lhd Column, c_2 \lhd Column
\end{align*}
\begin{align*}
	\mathbf{\cong} : ForeignKey \times ForeignKey \rightarrow Boolean \\
	f_1 \cong f_2 = \begin{cases}
 		\inference{name(f_1) = name(f_2) \\ \wedge reference(f_1) = reference(f_2) \\ \wedge constraints(f_1) = comstraints(f_2}{true}
 		\\ \\
		false
 \end{cases}\\
 f_1 \lhd ForeignKey, f_2 \lhd ForeignKey
\end{align*}

\paragraph{Find pair} The function returns a pair corresponding to given column (or key) name in concrete row.
\begin{align*}
\mathbf{findPair} : Label \times Row \rightarrow Pair \\
	findPair(l, r) = \begin{cases}
		\inference{\exists p \in pairs(r) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
		l \in Label, r \lhd Row
\end{align*}
\subsubsection{Database Schema Creation}
\paragraph{Add Column} A new column can be add into a table if there are no name collisions. If the column has the $NOTNULL$ constraint, then the default value has to be added to all rows in the table.
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		%\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \not \in constraints(c) }{columns(t) \cup c}\\ \\
		(\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \in constraints(c) }{columns(t) \cup c})\\ \bullet addValueToTable(l, [name(c), defVal(c)]) 
%		\\ \\ Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}

\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		%\inference{|t| > 1}{d \bullet addTable(head(t)) \bullet addTable(tail(t))}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge \forall f \in foreignKeys(t) : \\ s\exists e \in tables(d) \wedge name(e) = reference(f) }{tables(d) \cup t}
% \\ \\ 	Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}

\subsubsection{Database Schema Alternation}
\paragraph{Alter Column Name}
\begin{align*}
\mathbf{alterColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) =
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
 \\
l_c, l_o, l_n \in Label, d \lhd Database
\end{align*}

@@

\subsubsection{Data Manipulation}
\paragraph{Insert row} The function allow to insert a row(s) into a database.
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			%\inference{|r|>1}{d \bullet insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) \\ \wedge \forall e \in tableData(t, d) : pkValue(e) \neq pkValue(r)}{data(d) \cup r} 
		 	%\\ \\	 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}
